% =============================================================================

\paragraph{Hardware.}
\label{sec:ise:eval:hw}

Each ISE variant was integrated into the two host cores 
described in \REFSEC{sec:ise:imp}.
The variants which assume  $\RVXLEN = 32$
(\ISE{1}, \ISE{2}, \ISE{3}, and \ISE{5}) 
were evaluated
on {\em both} the
$32$-bit \CORE{2} core
{\em  and} the
$32$-bit \CORE{1} core;
the variant  which assumes $\RVXLEN = 64$
(\ISE{4})
was  evaluated
on {\em only} the
$64$-bit \CORE{1} core.
For \ISE{1}, \ISE{2} and \ISE{5} a trade-off
between latency and area exists. 
Each such case is considered through two optimisation goals:
the (A)rea    goal
instantiates $1$ S-box   and has a $n$-cycle execution latency,
whereas
the (L)atency goal
instantiates $4$ S-boxes and has a $1$-cycle execution latency.
We focus on ASIC implementations (rather than FPGA implementations)
because this is the more relevant metric to the industrial (rather than
academic) RISC-V community.

\REFTAB{tab:eval:hw:encdec}
shows the separated cost of the standalone ISE logic and the
combined cost of the core and integrated ISE.
Numbers highlighted in {\bf bold} are the best result for each metric.
The {\em Baseline} rows indicate the metrics for the the
{\em unmodified} host CPU cores.
We use the open source Yosys~\cite{yosys} synthesis tool (v0.9+1706)
with default settings
to provide post-synthesis (as opposed to post-layout) circuit area in the
form of NAND2 gate equivalents
(ISE Area, Tables \ref{tab:eval:hw:encdec} and \ref{tab:eval:hw:dec})
and circuit depths
in the form of gate delays
(ISE Latency, Tables \ref{tab:eval:hw:encdec} and \ref{tab:eval:hw:dec}).
While more abstract than providing exact area and
frequency results for a particular ASIC standard cell library, it is
much easier to reproduce\footnote{
Especially
for researchers lacking expensive commercial
synthesis tools and process design kits.
} while still providing meaningful results.
This methodology has also been used for other RISC-V standard extension
proposals, namely the bit-manipulation extension~\cite[Section 3.1, Page 54]{riscv:bitmanip:draft}.
We found 
that none of the ISEs affected the critical gate delay path of either the
\CORE{2} or \CORE{1} core.
These were $97$ for the 32-bit \CORE{2} core and $231$ and $167$ for the
$32$ and $64$-bit \CORE{1} core respectivley\footnote{
We are unable to explain why the gate delay path should be longer for
the $32$-bit \CORE{2} core than the $64$-bit variant without a deep dive
into the micro-architecture. We suspect it is an artifact of the sheer
configurability (rather than optimality) of the \CORE{1} core.
}.
Considering each ISE as implemented on the \CORE{1} core, we note the 
overhead wrt. area is marginal: this stems from the fact that the 
baseline area of \CORE{1} includes the data and instruction caches.

In
\REFTAB{tab:eval:hw:dec}
we consider the hardware costs when only {\em encryption} instructions are
implemented.
This is relevant to systems which only care about certain block cipher
modes of operation, such as Galos/Counter-mode.
We discuss this further in \REFSEC{sec:gcm}.

\begin{table}[p]
\centering
\begin{tabular}{|c|l|rr|r|r|}
\hline
  \multicolumn{1}{|c|}{ISA}
& \multicolumn{1}{ c|}{Variant}
& \multicolumn{1}{ c|}{             ISE}
& \multicolumn{1}{ c|}{       ISE      }
& \multicolumn{1}{ c|}{\CORE{2}     CPU}
& \multicolumn{1}{ c|}{\CORE{1}     CPU}
\\
& \multicolumn{1}{ c|}{/ Goal       }
& \multicolumn{1}{ c|}{Area         }
& \multicolumn{1}{ c|}{Latency      }
& \multicolumn{1}{ c|}{$+$ ISE area }
& \multicolumn{1}{ c|}{$+$ ISE area }
\\
\hline
\hline
 RV32IMC & Baseline    &              &            &       37325  ($1.00\times$) &       3501576 ($1.000\times$) \\
 RV32IMC & \ISE{1} (L) &        3514  & \bftab 18  &       41746  ($1.12\times$) &       3508448 ($1.002\times$) \\
 RV32IMC & \ISE{1} (A) &        2195  &        21  &       40171  ($1.08\times$) &       3506995 ($1.002\times$) \\
 RV32IMC & \ISE{2} (L) &        3574  &        19  &       41132  ($1.10\times$) &       3508946 ($1.002\times$) \\
 RV32IMC & \ISE{2} (A) &        1355  &        21  &       38777  ($1.04\times$) &\bftab 3506591 ($1.001\times$) \\
 RV32IMC & \ISE{3}     & \bftab 1149  &        30  &\bftab 38546  ($1.03\times$) &       3506761 ($1.001\times$) \\
 RV32IMC & \ISE{5} (L) &        4172  &        21  &       42035  ($1.13\times$) &       3510055 ($1.002\times$) \\
 RV32IMC & \ISE{5} (A) &        1726  &        23  &       39144  ($1.05\times$) &       3507755 ($1.002\times$) \\
\hline
\hline
 RV64IMC & Baseline &          &          &  N/A  & 3717607 (1.000$\times$) \\
 RV64IMC & \ISE{4}  &     8226 &       28 &  N/A  & 3733786 (1.004$\times$) \\
\hline
\end{tabular}
\caption{
Hardware metrics for each ISE variant with encrypt and decrypt instructions.
Area is measured in NAND2 gate equivalents and latency in gate delays.
}
\label{tab:eval:hw:encdec}
\end{table}
% =============================================================================

\paragraph{Software.}
\label{sec:ise:eval:sw}

We evaluated each ISE variant by implementing the AES-128
$\ALG{Enc}$,
$\ALG{Dec}$
{\em plus}
$\ALG{Enc-KeyExp}$
and
$\ALG{Dec-KeyExp}$.
We use our own implementation of a {\em non}-ISE 
T-table based
implementation as a baseline.
The variants which assume  $\RVXLEN = 32$
(\ISE{1}, \ISE{2},     \ISE{3}, and \ISE{5})
used a    rolled strategy wrt. loops:
 \ISE{1}, \ISE{2},              and \ISE{5}
used  $1$ round  per iteration,
whereas
                       \ISE{3}
used $2$ rounds per iteration
to avoid needless register move operations.
The variant  which assumes $\RVXLEN = 64$
(\ISE{4})
used an unrolled strategy.
In all cases the state is naturally aligned,\footnote{%
RISC-V does not mandate support for misaligned loads and stores, so
aligning the state this way ensures the best performance across all
cores.
} meaning any input (resp. output) can be loaded (resp. stored) 
using 
$4$ \VERB{lw} instructions on a $32$-bit core
or
$2$ \VERB{ld} instructions on a $64$-bit core.

\REFTAB{tab:eval:sw:size} 
records the
memory footprint (i.e., code footprint and static data footprint)
of each software implementation.
Again, numbers highlighted in {\bf bold} are the best result for each metric.
Where an entry for
\ALG{Dec-KeyExp}
is     zero, this implies that
$\ALG{Enc-KeyExp} =    \ALG{Dec-KeyExp}$
so there is no overhead.
Where an entry for
\ALG{Dec-KeyExp}
is non-zero, this implies that
$\ALG{Enc-KeyExp} \neq \ALG{Dec-KeyExp}$,
and the equivalent inverse cipher construction~\cite[Section 5.3.5]{FIPS:197}
is used.
This allows $\ALG{Dec-KeyExp}$ to
call $\ALG{Enc-KeyExp}$,
then
perform some additional post processing,
with the quoted footprint therefore reflecting the latter only.  
\REFTAB{tab:eval:sw:perf:2}
and
\REFTAB{tab:eval:sw:perf:1}
record
instruction (i.e., iret) and cycle counts
of each implementation,
as executed on the \CORE{2} and \CORE{1} cores respectively.


\begin{table}[p]
\centering
\begin{tabular}{|c|c|r|r|r|r|r|}
\hline
  \multicolumn{1}{|c|}{ISA}
& \multicolumn{1}{ c|}{Variant}
& \multicolumn{1}{ c|}{$\ALG{Enc}$}
& \multicolumn{1}{ c|}{$\ALG{Dec}$}
& \multicolumn{1}{ c|}{$\ALG{Enc-KeyExp}$}
& \multicolumn{1}{ c|}{$\ALG{Dec-KeyExp}$}
& \multicolumn{1}{ c|}{.data} 
\\
\hline
\hline
%RV32IMC & Byte    &            &           &      312 &        0 &  522 \\
 RV32IMC & T-table &       804  &       804 &      154 &      174 & 5120 \\
 RV32IMC & \ISE{1} &       424  &       424 &\bftab 68 & \bftab 0 &\bftab 10 \\
 RV32IMC & \ISE{2} &\bftab 234  &\bftab 238 &\bftab 68 &       62 &\bftab 10 \\
 RV32IMC & \ISE{3} &       290  &       290 &       86 &       64 &\bftab 10 \\
 RV32IMC & \ISE{5} &       266  &       278 &      290 & \bftab 0 &\bftab 10 \\
\hline
 RV64IMC & \ISE{4} &       268  &       268 &      168 &      100 &    0 \\
\hline
\end{tabular}
\caption{Software memory footprint measured in bytes for each ISE variant.}
\label{tab:eval:sw:size}
\end{table}

\begin{table}[p]
\centering
\begin{tabular}{|c|l|rr|rr|rr|rr|}
\hline
  \multicolumn{1}{|c|}{ISA}
& \multicolumn{1}{ c|}{Variant}
& \multicolumn{2}{ c|}{$\ALG{Enc}$}
& \multicolumn{2}{ c|}{$\ALG{Dec}$}
& \multicolumn{2}{ c|}{$\ALG{Enc-KeyExp}$}
& \multicolumn{2}{ c|}{$\ALG{Dec-KeyExp}$}
\\
\cline{3-10}
& / Goal
& \multicolumn{1}{ c|}{iret}
& \multicolumn{1}{ c|}{cycles}
& \multicolumn{1}{ c|}{iret}
& \multicolumn{1}{ c|}{cycles}
& \multicolumn{1}{ c|}{iret}
& \multicolumn{1}{ c|}{cycles}
& \multicolumn{1}{ c|}{iret}
& \multicolumn{1}{ c|}{cycles}
\\
\hline
\hline
%RV32IMC & Byte        &            &            &            &            &            &            &            &            \\
 RV32IMC & T-table     &          938 &         1016 &          938 &         1037&          430 &          515 &         1711 &         2307 \\ 
 RV32IMC & \ISE{1} (L) &          512 &          575 &          512 &          576& \bftab   198 & \bftab   302 & \bftab   204 & \bftab   321 \\
 RV32IMC & \ISE{1} (A) &          512 &          735 &          512 &          736& \bftab   198 &          342 & \bftab   204 &          361 \\
 RV32IMC & \ISE{2} (L) & \bftab   215 & \bftab   274 & \bftab   216 & \bftab   285& \bftab   198 & \bftab   302 &          335 &          615 \\
 RV32IMC & \ISE{2} (A) & \bftab   215 &          501 & \bftab   216 &          522& \bftab   198 &          332 &          335 &          753 \\
 RV32IMC & \ISE{3}     &          238 &          291 &          238 &          286&          219 &          312 &          659 &         1118 \\
 RV32IMC & \ISE{5} (L) &          227 &          294 &          227 &          291&          332 &          449 &          338 &          468 \\
 RV32IMC & \ISE{5} (A) &          227 &          554 &          227 &          532&          332 &          479 &          338 &          498 \\
\hline
\end{tabular}                
\caption{Execution metrics for each ISE variant on the \CORE{2} core.  Note that the $64$-bit \ISE{4} is absent, since there is no $64$-bit \CORE{2} core.}
\label{tab:eval:sw:perf:2}
\end{table}

\begin{table}[p]
\centering
\begin{tabular}{|c|l|rr|rr|rr|rr|}
\hline
  \multicolumn{1}{|c|}{ISA}
& \multicolumn{1}{ c|}{Variant}
& \multicolumn{2}{ c|}{$\ALG{Enc}$}
& \multicolumn{2}{ c|}{$\ALG{Dec}$}
& \multicolumn{2}{ c|}{$\ALG{Enc-KeyExp}$}
& \multicolumn{2}{ c|}{$\ALG{Dec-KeyExp}$}
\\
\cline{3-10}
& / Goal
& \multicolumn{1}{ c|}{iret}
& \multicolumn{1}{ c|}{cycles}
& \multicolumn{1}{ c|}{iret}
& \multicolumn{1}{ c|}{cycles}
& \multicolumn{1}{ c|}{iret}
& \multicolumn{1}{ c|}{cycles}
& \multicolumn{1}{ c|}{iret}
& \multicolumn{1}{ c|}{cycles}
\\
\hline
\hline
%RV32IMC & Byte        &            &            &            &            &            &            &            &            \\
 RV32IMC & T-table     &       934  &      1338  &       934  &      1003  &       430  &       569  &      1711  &      2167  \\
 RV32IMC & \ISE{1} (L) &       513  &       659  &       513  &       613  &\bftab 199  &       268  &\bftab 200  &\bftab 270  \\
 RV32IMC & \ISE{1} (A) &       513  &       791  &       513  &       725  &\bftab 199  &       308  &\bftab 200  &       310  \\
 RV32IMC & \ISE{2} (L) &\bftab 216  &\bftab 351  &\bftab 217  &       354  &\bftab 199  &\bftab 263  &       336  &       496  \\
 RV32IMC & \ISE{2} (A) &\bftab 216  &       503  &\bftab 217  &       534  &\bftab 199  &       293  &       336  &       637  \\
 RV32IMC & \ISE{3}     &       239  &       396  &       239  &       410  &       220  &       462  &       660  &      2420  \\
 RV32IMC & \ISE{5} (L) &       228  &       366  &       228  &\bftab 322  &       333  &       405  &       334  &       404  \\
 RV32IMC & \ISE{5} (A) &       228  &       536  &       228  &       546  &       333  &       438  &       334  &       434  \\
\hline
 RV64IMC & T-table     &       934  &      1086  &       934  &      1015  &       431  &       479  &      1712  &      1995  \\
 RV64IMC & \ISE{4}     &        76  &       115  &        76  &       133  &        61  &       199  &       131  &       286  \\
\hline
\end{tabular}
\caption{Execution metrics for each ISE variant on the \CORE{1} core.  Note that the $64$-bit \ISE{4} uses the $64$-bit \CORE{1} core; all others use the $32$-bit \CORE{1} core.}
\label{tab:eval:sw:perf:1}
\end{table}
% =============================================================================

\paragraph{Discussion.}
\label{sec:ise:eval:discuss}

\REFTAB{tab:eval:hw:encdec}
demonstrates that all ISE variants
imply a modest area overhead relative to their host core.
For the RV32 \CORE{1} the area overhead of a synthesised \CORE{1} Tile with
caches was less than $1\%$ in all cases.
For the \CORE{2}, the area overhead ranged between
$13\%$ (\ISE{5} (L))
and
$3\%$ (\ISE{3}).
\REFTAB{tab:eval:sw:size}
shows all ISE variants
having similarly small memory footprints in terms of both instruction code and
data.
Beyond this, and per 
\REFSEC{sec:ise:design},
the primary metric of interest is efficiency in terms of
the latency-area product.
This metric draws on data from
\REFTAB{tab:eval:hw:encdec}
plus either
\REFTAB{tab:eval:sw:perf:2}
or
\REFTAB{tab:eval:sw:perf:1}
for the \CORE{2} or \CORE{1} core respectively.
We note the small difference in instruction count in some cases between the
cores.
This is due to slightly different compiler behaviour at the mesured
function call sites in each core: the \CORE{1} core saves an extra
register to the stack.
We deliberately omit the area of the host core from this calculation, as this
fixed overhead dominates the final value and detracts from the comparison
between ISEs themselves.

\REFTAB{tab:eval:results} 
captures the results for the \CORE{1} core, although the same conclusion can 
be drawn for the \CORE{2} core.  Qualitatively, we place more of a weight on 
Encryption (\ALG{Enc})
and 
Decryption (\ALG{Dec})
vs.
Encryption Key Expansion (\ALG{Enc-KeyExp})
and 
Decryption Key Expansion (\ALG{Dec-KeyExp}),
because
typically many \ALG{Enc} or \ALG{Dec} operations are performed per
\ALG{KeyExp}.

For a $32$-bit core, our conclusion is that
\ISE{3} 
is the best option.
Despite not being the fastest (by a small margin), it is the most efficient,
and simplest to implement.
The area optimised \ISE{2} implementation sometimes comes close in
efficiency, but requires a more complex multi-cycle implementation
in this case.
We note that \ISE{3} has relatively poor performance for the decryption
key schedule.
This is because it uses the Equivalent Inverse Cipher representation,
and must first create an {\em encryption} orientated key schedule, before
applying the Inverse \AESFUNC{MixColumns} transform to each word in the key
schedule.
Each word requires $8$ instructions to apply {\em only} the Inverse \AESFUNC{MixColumns} 
transform. We believe this is reasonable, as one typically performs
many block decryptions per key schedule operation.
We also note that for the common AES-GCM usecase, decryption functionality
is not necessary. We discuss this further in \REFSEC{sec:gcm}.
Compared to past work, our implementation of \ISE{3} is
slightly smaller than its original description in \cite{Saarinen:20}:
$1149$ v.s. $1240$ gates.
\cite{Saarinen:20} estimates a $5\times$ performance improvement, which is
slightly better than our measured $4\times$ improvement, though this is
dependant on relative memory access latencies.
We would expect this improvement to increase in systems which store T-tables
in (relatively) high latency flash memory.
\ISE{3} performs considerably better than \cite{TilGroSze:05},
which achieves only a $2\times$ speedup in the best case.
We note that despite needing the same number of instructions per round
as \ISE{2} (based on \cite{TilGroSze:05}), our \ISE{5} design suffers in terms
of performance.
This is due to the conversion between quadrant-packed and column-packed
representations.

For a $64$-bit core,
\ISE{4} 
is the best option, which is somewhat obvious because it specifically makes
use of the wider data-path.
It is $10\times$ faster to perform a block encryption than a baseline T-table
implementation targeting a 64-bit base RISC-V architecture.
With reference to
\REFTAB{tab:eval:sw:perf:1}, 
note that the number of cycles per instruction executed is relatively high.
This fact stems from use of the ROCC interface, in that forwarding of the 
result from an ISE instruction (that uses the ROCC) incurs an overhead vs. 
an ISE instruction; fine-grained integration of the AES-FU could therefore
incrementally improve the results.

We believe it is sensible to standardise different ISEs for the
RV32 and RV64 base ISAs.
This allows each ISE design to better suit the constraints of each
base ISA.
In the RV32 case, this acknowledges that such cores will most often
appear in resource-constrained, embedded or IoT class devices.
Hence, the most efficient ISE design is appropriate.
For necessarily larger RV64-based designs, it makes sense to take advantage
of the wider data-path, and acknowledge that these are more likely to
be application class cores. Hence, they will place a higher value
on performance than area-efficiency.

% -----------------------------------------------------------------------------


% -----------------------------------------------------------------------------

\begin{table}[p]
\centering
\begin{tabular}{|c|l|rr|rr|rr|rr|}
\hline
  \multicolumn{1}{|c|}{ISA}
& \multicolumn{1}{ c|}{Variant}
& \multicolumn{2}{ c|}{$\ALG{Enc}$}
& \multicolumn{2}{ c|}{$\ALG{Dec}$}
& \multicolumn{2}{ c|}{$\ALG{EncKeyExp}$}
& \multicolumn{2}{ c|}{$\ALG{DecKeyExp}$}
\\
\cline{3-10}
& / Goal
& \multicolumn{1}{ c|}{iret}
& \multicolumn{1}{ c|}{cycles}
& \multicolumn{1}{ c|}{iret}
& \multicolumn{1}{ c|}{cycles}
& \multicolumn{1}{ c|}{iret}
& \multicolumn{1}{ c|}{cycles}
& \multicolumn{1}{ c|}{iret}
& \multicolumn{1}{ c|}{cycles}
\\
\hline
\hline
RV32IMC & \ISE{1} (L) &        1.00 &           1.00 &           1.00 &           1.00 &           0.50 &        0.57 &        0.51 &        0.51 \\
RV32IMC & \ISE{1} (A) &        0.62 &           0.80 &           0.62 &           0.80 &           0.31 &        0.40 & \bftab 0.32 & \bftab 0.36 \\
RV32IMC & \ISE{2} (L) &        0.43 &           0.48 &           0.43 &           0.50 &           0.51 &        0.58 &        0.85 &        1.00 \\
RV32IMC & \ISE{2} (A) &        0.16 &           0.34 &           0.16 &           0.35 &           0.19 &        0.24 & \bftab 0.32 &        0.46 \\
RV32IMC & \ISE{3}     & \bftab 0.15 & \bftab    0.17 & \bftab    0.15 & \bftab    0.16 & \bftab    0.18 & \bftab 0.19 &        0.54 &        0.58 \\
RV32IMC & \ISE{5} (L) &        0.53 &           0.61 &           0.53 &           0.60 &           1.00 &        1.00 &        1.00 &        0.89 \\
RV32IMC & \ISE{5} (A) &        0.22 &           0.47 &           0.22 &           0.45 &           0.41 &        0.44 &        0.41 &        0.39 \\
\hline
RV64IMC & \ISE{4}     &       0.266 &       0.402 &       0.266 &       0.465 &       0.213 &       0.696 &       0.458 &       1.000  \\
\hline
\end{tabular}
\caption{Comparison of performance/area product. Each value is normalised to the largest product per column. The RV64IMC row is not normalised as there is no comparison point.}
\label{tab:eval:results}
\end{table}

% =============================================================================
