% =============================================================================

\subsection{Hardware Evaluation}
\label{sec:eval:sw}

Each ISE variant was evaluated on the host cores
described in \REFSEC{sec:design}.
The 32-bit designs (V1,V2,V3,V5) were implemented on both the
\CORE{1} and \CORE{2} cores.
The 64-bit design (v4) was only evaluated on the 64-bit configuration
of the \CORE{1} core.
Table \ref{tab:eval:hw} shows the hardware implementation costs.

For variants 1, 2 and 5, two implementations are evaluated.
The {\em Size} optimised implementations instantiate only a single
Forward/Inverse SBox and MixColumns circuit and take multiple cycles
to produce a result.
The {\em Latency} optimised implementations instantiate $4$ SBox and
MixColumn circuits to produce their results in a single processor 
clock cycle.

The {\em Size} columns of Table \ref{tab:eval:hw} 
record the number of NAND2 equivalent gates of each variant,
instantiated independently from any wider system.
The LTP column gives the Longest Topological Path of the synthesised
functional unit circuit from input to combinatorial output.
The \CORE{2} Size column gives the size in NAND2 equivalent gates of the
\CORE{2}, with the various AES functional units integrated.
The ``Baseline" row gives the size of the core without any of the
ISEs integrated.
We found that none of the proposed ISEs affected the critical
path of the \CORE{2} core.

% ------------------------------------------------------------

\begin{table}
\centering
\begin{tabular}{lrrrr}
Variant     & ISE Size & ISE LTP & \CORE{2} Size   & \CORE{1} Size \\ \hline
Baseline    & -        & -       & 37375 ($1.00$x) &               \\
V1 (Latency)& 3472     & 19      & 41723 ($1.12$x) &               \\
V2 (Latency)& 3547     & 19      & 41199 ($1.10$x) &               \\
V5 (Latency)& 4121     & 22      & 42070 ($1.13$x) &               \\
V3          & 1157     & 30      & 38610 ($1.03$x) &               \\
V4          &          &         & N/A             & {\bf TODO}    \\
V1 (Size)   & 2174     & 22      & 40161 ($1.07$x) &               \\
V2 (Size)   & 1381     & 21      & 38885 ($1.04$x) &               \\
V5 (Size)   & 1927     & 23      & 39251 ($1.05$x) &               \\
\end{tabular}
\caption{
Hardware implementation costs of each ISE.
The ISE columns list the size and depth of the ISE verilog hardware
descriptions when instantiated in isolation.
The \CORE{2} and \CORE{1} columns give the size (in NAND2 equivalent gates)
of the ISEs when integrated into each core.
Gate counts and topological path lengths are obtained using the
Yosys\cite{yosys} tool suite.
}
\label{tab:eval:hw}
\end{table}



% =============================================================================
