\documentclass[preprint]{iacrtrans}

\usepackage{paper}

\title{An exploration of ISEs for AES on RISC-V}
\keywords{ISE, AES, RISC-V}

\ifbool{anonymous}{%
\author{}
\institute{}
}{%
\author{
Andy Glew\inst{1}                   \and
Ben Marshall\inst{2}                \and
G. Richard Newell\inst{3}           \and
Dan Page\inst{2}                    \and
Markku-Juhani O. Saarinen\inst{4}   \and
Barry Spinney\inst{5}               \and
Claire Wolf\inst{6}
}
\institute{
SiFive, Inc. \\ \email{andy.glew@sifive.com}
\and
Department of Computer Science, University of Bristol \\ \email{{ben.marshall,daniel.page}@bristol.ac.uk}
\and
Microchip Technology Inc., USA \\ \email{richard.newell@microchip.com}
\and
PQShield, UK \\ \email{mjos@pqshield.com}
\and
Nvidia Corporation \\ \email{first.last@nvidia.com}
\and
Symbiotic EDA \\ \email{claire@symbioticeda.com}
}
}%

\begin{document}

% =============================================================================

\maketitle

\begin{abstract}
Secure, efficient execution of AES is an essential requirement for most
computing platforms, often motivating inclusion of dedicated
Instruction Set Extensions (ISEs).
RISC-V is a (relatively) new ISA that lacks such a standardised ISE.
We survey the state-of-the-art industrial and academic ISEs for AES,
implement and evaluate five different ISEs, one of which is novel,
and make recommendations for standardisation.
We consider the side-channel security implications of the ISE designs,
demonstrating how an implementation of one candidate ISE can be hardened
against DPA-style attacks.
We also explore how the proposed standard Bit-manipulation extension
to RISC-V can be harnessed for efficient implementation of AES-GCM.
Our work supports the ongoing RISC-V cryptography extension standardisation
process.
\end{abstract}

% =============================================================================

\section{Introduction}
\label{sec:intro}
\input{tex/intro.tex}
\input{tex/intro-riscv.tex}
\input{tex/intro-remit.tex}

% =============================================================================

\section{Background}
\label{sec:bg}

FIPS-197~\cite{FIPS:197} represents the definitive specification
of AES. 
An overview of related design rationale is offered in ~\cite{DaeRij:02}.
We endeavour to follow the notation set out in ~\cite{FIPS:197}
in referencing specific parts of AES functionality.

% -----------------------------------------------------------------------------

%\subsection{AES specification}
%\label{sec:bg:aes_spec}
%\input{tex/bg-aes_spec.tex}

\subsection{AES implementation}
\label{sec:bg:aes_impl}

\paragraph{Representation.}
\label{sec:bg:aes_impl_rep}
\input{tex/bg-aes_impl_rep.tex}

\paragraph{Hardware-only implementations.}
\label{sec:bg:aes_impl_hw}
\input{tex/bg-aes_impl_hw.tex}
\paragraph{Software-only implementations.}
\label{sec:bg:aes_impl_sw}
\input{tex/bg-aes_impl_sw.tex}
\subsection{Existing AES ISEs}
\label{sec:bg:aes_impl_ise}
\input{tex/bg-aes_impl_ise.tex}

\paragraph{Security}
\label{sec:bg:aes_impl_sec}
\input{tex/bg-aes_impl_sec.tex}

% =============================================================================

\section{Exploring AES ISEs for RISC-V}
\label{sec:ise}

% -----------------------------------------------------------------------------

\label{sec:ise:design}

\input{tex/body-ise_intro.tex}

\subsection{Variant 1 (\ISE{1}): \AESFUNC{SubBytes} $+$ \AESFUNC{MixColumn} $+$ explicit \AESFUNC{ShiftRows}}
\label{sec:ise:design:v1}
\input{tex/body-ise_design_v1.tex}
\subsection{Variant 2 (\ISE{2}): \AESFUNC{SubBytes} $+$ \AESFUNC{MixColumn} $+$ implicit \AESFUNC{ShiftRows}}
\label{sec:ise:design:v2}
\input{tex/body-ise_design_v2.tex}
\subsection{Variant 3 (\ISE{3}): hardware-assisted T-tables}
\label{sec:ise:design:v3}
\input{tex/body-ise_design_v3.tex}
\subsection{Variant 4 (\ISE{4}): $64$-bit data-path}
\label{sec:ise:design:v4}
\input{tex/body-ise_design_v4.tex}
\subsection{Variant 5 (\ISE{5}): quadrant-packed}
\label{sec:ise:design:v5}
\input{tex/body-ise_design_v5.tex}

% -----------------------------------------------------------------------------

\subsection{Implementation}
\label{sec:ise:imp}
\input{tex/body-ise_implement.tex}

% -----------------------------------------------------------------------------

\subsection{Evaluation}

\paragraph{Hardware}
\label{sec:ise:eval:hw}
\input{tex/body-ise_eval_hw.tex}

\paragraph{Software}
\label{sec:ise:eval:sw}
\input{tex/body-ise_eval_sw.tex}

\paragraph{Discussion}
\label{sec:ise:eval:discuss}
\input{tex/body-ise_eval_discuss.tex}

% =============================================================================

\section{Using ISEs to implement AES-GCM}
\label{sec:gcm}
%\input{tex/body-modes.tex}
\input{tex/body-modes-gcm.tex}

% =============================================================================

\section{Hardening an AES ISE against DPA attack}
\label{sec:sca}
\input{tex/body-sca.tex}

% =============================================================================

\section{Conclusion}
\label{sec:outro}
\input{tex/outro.tex}

% ============================================================================

\ifbool{anonymous}{}{%
\section*{Acknowledgements}

This work was undertaken as part of the ongoing standardisation of
RISC-V. We are grateful to {\em all} members of
the Cryptographic Extensions Task Group
who contributed to related discussions.
The opinions expressed in this paper are the author's alone, not
of their respective employers or the RISC-V International Foundation.
The RISC-V cryptography extension is in the process of being standardised
at the time of writing. The purpose of this work is to support that process.

We would like to thank the anonymous reviewers for their helpful and 
constructive comments.

This work has been supported in part by EPSRC via grant EP/R012288/1, 
under the RISE (\url{http://www.ukrise.org}) programme.
}%

% =============================================================================

\bibliographystyle{alpha}
\bibliography{paper}

% =============================================================================

\appendix

\clearpage
\section{\ISE{1}:       additional technical detail}
\label{sec:pseudo:v1}
\input{tex/appx-design_v1.tex}
\clearpage
\section{\ISE{2}:       additional technical detail}
\label{sec:pseudo:v2}
\input{tex/appx-design_v2.tex}
\clearpage
\section{\ISE{3}:       additional technical detail}
\label{sec:pseudo:v3}
\input{tex/appx-design_v3.tex}
\clearpage
\section{\ISE{4}:       additional technical detail}
\label{sec:pseudo:v4}
\input{tex/appx-design_v4.tex}
\clearpage
\section{\ISE{5}:       additional technical detail}
\label{sec:pseudo:v5}
\input{tex/appx-design_v5.tex}

\clearpage
\section{\mbox{\CORE{2}} core: additional technical detail}
\label{sec:core:2}
\input{tex/appx-core_2.tex}
\clearpage
\section{\mbox{\CORE{1}} core: additional technical detail}
\label{sec:core:1}
\input{tex/appx-core_1.tex}

%\clearpage
%\section{Additional algorithms}
%\label{sec:alg}
%\input{tex/appx-alg.tex}

% =============================================================================

\end{document}
